---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# ECPIN

<!-- badges: start -->
<!-- badges: end -->

The current project aims to automatically analyze environmental chemcials binding to proteins through the APNA method

## Installation

You can install the the development version from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("huiUofT/ECPIN")
```
## Example

This is a basic example which shows you how to solve a common problem:

Step 1: loading packages
```{r, message=FALSE, warning=FALSE}
  rm(list=ls())
  library(xcms)
  library(eCPIN)
  library(devtools)
  library(xlsx)
  library(ChemmineR)
  library(rcdk)
  library(MassSpecWavelet)
  library(Rcpp)
  library(RcppArmadillo)
  library(isopat)
  library(readxl)
  library(ggplot2)
  library(tidyverse)
  library(dplyr)
  library(ggrepel)
  library(openxlsx)
  data(iso_list)
iso.otheratom<-rbind(1:6,1:6)###save the isotopic distribution of other atoms
iso.otheratom[1,1]<-iso_list[4,3]-iso_list[3,3]##mz difference of C
iso.otheratom[2,1]<-iso_list[4,4]/iso_list[3,4]##relative abundance of C
iso.otheratom[1,2]<-iso_list[2,3]-iso_list[1,3]##mz difference of H
iso.otheratom[2,2]<-iso_list[2,4]/iso_list[1,4]##relative abundance of H
iso.otheratom[1,3]<-iso_list[6,3]-iso_list[5,3]##mz difference of N
iso.otheratom[2,3]<-iso_list[6,4]/iso_list[5,4]##relative abundance of N
iso.otheratom[1,4]<-iso_list[9,3]-iso_list[7,3]##mz difference of O
iso.otheratom[2,4]<-iso_list[9,4]/iso_list[7,4]##relative abundance of O
iso.otheratom[1,5]<-0##mz difference of P
iso.otheratom[2,5]<-0##relative abundance of P
iso.otheratom[1,6]<-iso_list[12,3]-iso_list[10,3]##mz difference of S
iso.otheratom[2,6]<-iso_list[12,4]/iso_list[10,4]##relative abundance of S

#################Rcpp function to enhance computation efficiency###############
cppFunction(
  'NumericMatrix itercal(NumericMatrix numberset, NumericVector mz_list,double ppm, double mz, double mwoffset){
  NumericMatrix output(500,12);
  int i1;
  int i2;
  int i3;
  int i4;
  int i5;
  int i6;
  int i7;
  int i8;
  int i9;
  int i10;
  int i11;
  int i12; 
  int kk=0;
  double temp1;
  double temp2;
  double temp3;
  double temp4;
  double temp;
  double value=100;
  double RDBE;
  for (i1=numberset(0,0);i1<=numberset(0,1);i1++){
  for (i2=numberset(1,0);i2<=(numberset(1,1)<(i1*2+3)?numberset(1,1):(i1*2+3));i2++){
  for (i3=numberset(2,0);i3<=(numberset(2,1)<(i1*1.3)?numberset(2,1):(i1*1.3));i3++){
  for (i4=numberset(3,0);i4<=(numberset(3,1)<(i1*1.2+2)?numberset(3,1):(i1*1.2+2));i4++){
  for (i5=numberset(4,0);i5<=(numberset(4,1)<(i1*0.3)?numberset(4,1):(i1*0.3));i5++){
  for (i6=numberset(5,0);i6<=numberset(5,1);i6++){
  for (i7=numberset(6,0);i7<=numberset(6,1);i7++){
  temp1=i1*mz_list[0]+i2*mz_list[1]+i3*mz_list[2]+i4*mz_list[3]+i5*mz_list[4]+i6*mz_list[5]+i7*mz_list[6];
  if (temp1>(mz+1)){
  break;}
  for (i8=numberset(7,0);i8<=numberset(7,1);i8++){
  temp2=temp1+i8*mz_list[7];
  if (temp2>(mz+1)){
  break;}
  for (i9=numberset(8,0);i9<=numberset(8,1);i9++){
  temp3=temp2+i9*mz_list[8];
  if (temp3>(mz+1)){
  break;}
  for (i10=numberset(9,0);i10<=numberset(9,1);i10++){
  temp4=temp3+i10*mz_list[9];
  if (temp4>(mz+1)){
  break;}
  for (i12=0;i12<=1;i12++){
  for (i11=numberset(10,0);i11<=numberset(10,1);i11++){
  temp=temp4+i11*mz_list[10]+i12*mwoffset;
  value=abs(1000000*(mz-temp)/temp);
  RDBE=i1-0.5/(i2+i7+i8+i9+i10+i11+0.0001)+0.5/(i3+i5+0.5)+1;
  if (value<ppm&&RDBE>=0&&RDBE<40){
  NumericVector out=NumericVector::create(i1,i2,i3,i4,i5,i6,i7,i8,i9,i10,i11,value);
  output.row(kk)=out;
  kk=kk+1;
  if (kk>=500){return(output(Range(0,kk-1),Range(0,11)));}
  }
  if (temp>(mz+1)){break;}
  }}}}}}}}}}}}
  if (kk==0){
  kk=1;}
  return(output(Range(0,kk-1),Range(0,11)));
  }')
  
  #'set up the path
  path<-getwd()
  polarity<--1
  setwd(path)
```

Step 2: calibrating mass spectra
```{r, include= FALSE}
  #setwd('C:/Rprogram/ECPIN')
  #MassCalibration(polarity)
```

Step 3: extracting peak features
```{r, message=FALSE, warning=FALSE}
  setwd(path)

  #extracting peaks, intensity cutoff is 10^6, ppm = 2.5
  peaks.raw<-Peakextract(10^5,2.5)
  
  #'Extracting differential peaks
  #'data has to be organized like NR_XX_X, WT_XX_X, NR_DMSO_X
  Control<-'MeOH'
  peaks<-LigandFeatureID(peaks.raw,'neg',Control)#This is the function extracting differential peaks compared to controL
```
Step 4: identifying primary isotopic peaks, and remove redudant isotopic peaks
```{r, message=FALSE, warning=FALSE}
  #2 ppm is the mass tolerance to find isotopic peaks
  #'0.80 is the correlation coefficient to extract isotopic peaks
  #'3 is the intensity ratio cutoff between primary and other isotopic peaks
  peaks.iso<-FindIsotope(peaks,2,0.80,2)
```
Step 5: loading database, and do initial searching with exact mass
```{r, message=FALSE, warning=FALSE}
  setwd(path)
  Database<-read_excel("TSCA_TOX21_HMDB_CFMID.xlsx")#This is a combined database of TSCA, Tox21 and HMDB
  peaks.iso$mz<-peaks.iso$mz*(1+0*10^(-6))##simple mass calibration
  Library<-InitialSearch(peaks.iso,3,polarity,Database)#2ppm mass tolerance
  
  #delete redundant peaks
  mz<-NULL
  rt<-NULL
  indexsave<-NULL
  for (i in 1:nrow(Library)){
    index<-which(mz==Library$mz[i])
    if (length(index)==0){#new mz
      mz<-c(mz,Library$mz[i])
      rt<-c(rt,Library$rt[i])
      next
    }
    if (Library$rt[i]==rt[index[1]]){#retention time is identical
      indexsave<-c(indexsave,i)
      next
    }
    mz<-c(mz,Library$mz[i])
    rt<-c(rt,Library$rt[i])
  }
  if (length(indexsave)>0){
  Library<-Library[-indexsave,]}
  
  #'writing the initial searching
  write.table(Library,file='RXR_neg.csv',sep=',',row.names = FALSE)
```
Step 6: extracting fragments and isotopic peaks
```{r, message=FALSE, warning=FALSE}
  setwd(path)
  FragList<-GetFrag(Library,10^(-5),15)#extract fragments from DIA,10 is the width of DIA windows
  #FragList<-GetFrag_AIF(Library,10^(-5),polarity)#extract fragments from AIF
  setwd(path)
  
  #'Finding isotopic peaks with abundance bigger than 0.1%
  Isotope.Data<-IsotopeFind(Library,10^(-5),0.8)
```
Step 7: extracting adducts
```{r, message=FALSE, warning=FALSE}
  setwd(path)
  if (polarity==1){
  Adducts<-c('[M+H]+','[M+NH4]+')
  MW.adducts<-c(1.007825,18.03437)
  Adducts.Find<-AdductsFind(Library,MW.adducts,2*10^(-6),Adducts)
}

if (polarity==-1){
  Adducts<-c('[M-H]-','M+FA-H')
  MW.adducts<-c(-1.007825,44.99765)
  Adducts.Find<-AdductsFind(Library,MW.adducts,2*10^(-6),Adducts)
}
```
Step 8: calculating scores
```{r, message=FALSE, warning=FALSE}
  setwd(path)
  cutoff<-5000
  mylib.Target<-DatabaseSearching(Library,cutoff,polarity,Database,Isotope.Data,FragList,iso_list,'ms2true')
```
Step 9: weighted scores
```{r, message=FALSE, warning=FALSE}
  setwd(path)
  #' the path to save results
  path<-getwd()
  
  #' the path to save raw data
  path.data<-paste0(path,"/data")
  setwd(path.data)
  rawfiles<-list.files()
  xraw<-xcmsRaw(rawfiles[1],includeMSn = 1)
  precursor<-Preclist(xraw)
  weightK<-c(1,50,1,1,1,1)#weight for scores,the second for ms2 score
  setwd(path)
  mylib.score<-Finalscore(mylib.Target,weightK,precursor)
  Target.rt<-ScoreRT(mylib.score,Database)
  output<-Output(Target.rt,-1000)
  
  #'write MS2 fragments
  output$fragments<-rep(0,nrow(output))
  for (i in 1:nrow(output)){
    index<-which(FragList$precursor==output$mz[i])
    if (length(index)==0){next}
    fragment<-NULL
    for (j in 1:length(index)){
      fragment<-paste(c(fragment,';',c(FragList$mz[index[j]],',',FragList$intensity[index[j]])),collapse = '')
    }
    output$fragments[i]<-fragment
  }
  
  write.table(output,file='TRb_negative_final.csv',sep=',',row.names = FALSE)
  
  
```
Step 10: output
```{r, message=FALSE, warning=FALSE}
  setwd(path)
  mydata<- readxl::read_excel("Belugas.xlsx", sheet = "positive_clean")
  newdata<-NULL
  for (i in 1:nrow(mydata)){
    smiles<-strsplit(mydata$SMILES[i],';')
    smiles<-smiles[[1]]
    mserror<-strsplit(mydata$mserror[i],';')
    mserror<-mserror[[1]]
    MS2score<-strsplit(mydata$MS2score[i],';')
    MS2score<-MS2score[[1]]
    Adducts<-strsplit(mydata$Adducts[i],';')
    Adducts<-Adducts[[1]]
    for (j in 1:length(smiles)){
      temp<-mydata[i,]
      temp$SMILES<-smiles[j]
      temp$mserror<-mserror[j]
      temp$MS2score<-MS2score[j]
      temp$Adducts<-Adducts[j]
      newdata<-rbind(newdata,temp)
    }
  }
  
  #CALCULATING MOLECULAR DESCRIPTORS
descNames <- rcdk::get.desc.names(type = "all")
Descriptors<-matrix(rep(0,nrow(newdata)*287),nrow=nrow(newdata),ncol=287)
for (i in 1:nrow(newdata)){
  smiles<-newdata$SMILES[i]
  mols<-rcdk::parse.smiles(smiles)
  mols<-mols[[1]]
  descrps<-rcdk::eval.desc(mols, descNames)
  
  Descriptors[i,]<-as.numeric(descrps[1,])
}

Final<-data.frame(newdata,Descriptors)
colnames(Final)<-c(colnames(newdata),colnames(descrps))
  
#CALCULATING ADDUCTS
AdductsFind<-function(Library,xrawdata,index,mw.adducts,mz_tol,Adducts){
    score<-rep(0,length(index))
    for (j in 1:length(index)){
      temp<-index[j]
      mz<-Library$mz[temp]
      rt<-Library$rt[temp]*60
      mzrange<-xrawdata@mzrange
      minmz<-mzrange[1]
      maxmz<-mzrange[2]
      mzmin<-max(minmz,mz-mz*mz_tol)
      mzmax<-min(maxmz,mz+mz*mz_tol)
      rtrange<-xrawdata@scantime
      rtmin<-max(min(rtrange),rt-30)
      rtmax<-min(max(rtrange),rt+30)
      if (rtmax<0){next}
      peaks<-rawEIC(xrawdata,mzrange=cbind(mzmin,mzmax),rtrange=cbind(rtmin,rtmax))
      
      #'correlations to further confirm adducts
      native.peak<-peaks$intensity
      for (k in 1:length(Adducts)){
        mz0<-mz+mw.adducts[k]##mz of adducts
        if (Adducts[k]=='[2M-H]-'){
          mz0<-2*mz+1.007825
        }
        if (mz0>1000){next}
        mzmin<-max(minmz,mz0-mz0*mz_tol)
        mzmax<-min(maxmz,mz0+mz0*mz_tol)
        isotope.peak<-rawEIC(xrawdata,mzrange=cbind(mzmin,mzmax),rtrange=cbind(rtmin,rtmax))
        isotope.peak<-isotope.peak$intensity
        if (sd(isotope.peak)==0||sd(native.peak)==0){next}###0 values
        corr<-cor(native.peak,isotope.peak)
        if (corr>0.80){
        score[j]<-1
        }
      }}
  return(score)
}

Library<-Final
path.data<-paste0(path,"/data")
setwd(path.data)
msfiles<-list.files()

if (polarity==1){
  Adducts<-c('[M+H-H2O]+','[M+NH4]+','[M+CH3OH+H]+','[M+Na]+')
  mw.adducts<-c(-17.00274-1.007825,18.03437-1.007825,33.03404-1.007825,22.98977-1.007825)
}

if (polarity==-1){
  Adducts<-c('[M+Cl]-','[M+CH2O2-H]-','[2M-H]-')
  mw.adducts<-c(34.96885-1.007825,46.00548,0)
}


Library$ADDUCTS<-rep(0,nrow(Library))
for (i in 1:length(msfiles)){#12 plates
  index.plate<-which(Library$SampleID==i)
  if (length(index.plate)<1){next}
  xrawdata<-xcmsRaw(msfiles[i])
  index<-which(Library$Adducts[index.plate]=='[M+H]+')
  Adduct.score<-AdductsFind(Library,xrawdata,index.plate[index],mw.adducts,5*10^(-6),Adducts)
  Library$ADDUCTS[index.plate[index]]<-Adduct.score
}
setwd(path)
write.table(Library,file='POSITIVE.csv',sep=',',row.names = FALSE)
```
